Computational methods in landscape ecology

### 2024-08-29 Manuscript.qmd
Update Bib and manuscript

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/05-simulations.qmd >}}
{{< include Sections/08-software.qmd >}}

>>>>

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/03-connectivity.qmd >}}
{{< include Sections/05-simulations.qmd >}}
{{< include Sections/08-software.qmd >}}

### 2024-08-29 Manuscript.tex
Update Bib and manuscript

Despite the extensive use of the landscape mosaic model and the
landscape metrics based on it, major limitations have been identified
regarding both its ability to quantify the spatial structure of
landscapes and their relationship to ecological processes (Gustafson
1998; Li and Wu 2004; Dramstad 2009; Kupfer 2012; Lausch et al. 2015;
Frazier and Kedron 2017). In line with Dramstad (Dramstad 2009), we
review four major issues of landscape metrics, namely their sensitivity
to both spatial scale and thematic resolution, redundancy and
interpretation.

@@

Identifying the characteristic scale at which the organisms or processes
of interest operate is a critical first step towards establishing sound
relationships between pattern and process (Levin 1992; Wu 2007; Cushman
and McGarigal 2008). To that end, a key requirement it to understand the
sensitivity of landscape metrics to spatial scale in terms of both
resolution (grain) and extent.

A key concept to consider regarding grain size is the minimum mapping
unit (MMU), namely the smallest area entity that can be mapped and
analyzed as a discrete patch. Since the input maps should represent not
only the location of patches but also their shape, a grain size several
times smaller than the smallest patch (i.e., the MMU) in order to grasp
(even schematically) its shape. Even though the MMU is often assumed as
a natural characteristic of the dataset, it has an unsurprising impact
on the computed landscape metrics, especially in terms of the number of
patches and their size distribution (Saura 2002).

While the choice of grain is often constrained by the available data,
e.g., Landsat scenes at 30 m resolution, the selection of spatial extent
and the way that it is sampled varies largely depending on the
phenomenon of interest, e.g., from large transects of forests in natural
parks to small buffers along the urban-rural gradient. At the same time,
the choice of spatial extent and sampling strategy is often subject to
many arbitrary criteria that can largely influence the ultimate results.
Such an issue has long been recognized in physical geography as the
modifiable areal unit problem (MAUP) (Openshaw 1979, 1984). Approaches
to dealing with the MAUP in the context of landscape metrics include the
identification of individual entities that are ecologically meaningful
and not modifiable, the evaluation of ``optimal'' zoning systems that
maximize interzonal variations and minimize intrazonal variation and the
analysis of the sensitivity of the zoning configuration (Jelinski and Wu
1996; Duque et al. 2018; Fletcher Jr et al. 2023).

TODO: section about approaches to sample landscapes to compute metrics
and available software tools, e.g., zonal analysis with polygons (e.g.,
administrative limits), grids, transects or buffers along a landscape
gradient, moving windows (Bosch 2019; Hesselbarth et al. 2019; Huais
2024)?

The scale effects of landscape metrics can be assessed by computing
scalograms, i.e., the response curves of metrics to changing scales. For
instance, some metrics related to the number and density of patches and
edges show consistent linear and power-law scaling relationships to
changing grain across a wide diversity of simulated and real landscapes,
whereas other metrics exhibit less predictable responses (Turner et al.
1989; Saura and Martínez-Millán 2001; Wu et al. 2002; Wu 2004; Saura and
Castro 2007; Šı́mová and Gdulová 2012; Argañaraz and Entraigas 2014;
Frazier et al. 2021). Nonetheless, the effects of changing the sampling
extent can be less predictable than those of changing grain size (Wu
2004). For instance, in landscapes with small extent to grain ratio,
boundary effects are likely to dominate and truncate patches, resulting
in unreliable landscape metric computations. In order to address scale
issues and advance the science of scaling in landscape ecology, studies
should explicitly consider the concept of scope, namely the ratio of
extent to resolution (Schneider 2001). As suggested by Frazier (Frazier
2023), this can be done following three key steps, namely by considering
scope guidelines when designing experiments, explicitly reporting grain
and extent so that scope can be calculated allowing quantitative
comparability with other studes and undertaking comparisons and
replications based on scope. The latter is showcased in a proof of
concept example experiment, where landscapes from the same ecoregion are
sampled using different grain and extent and then the distributions of
the computed landscape metrics are visualized through ridgeline plots
colored by scope (Frazier 2023).

TODO: section about landscape metric rescaling (Saura and Castro 2007;
Argañaraz and Entraigas 2014; Frazier 2014, 2016; Frazier et al. 2021)?

TODO: mention hierarchichal patch dynamics (Wu 1999)?

TODO: sensitivity to thematic resolution (Huang et al. 2006; Bailey et
al. 2007; Buyantuyev and Wu 2007; Castilla et al. 2009).

@@

Many landscape ecology studies are based on the raster data model.
However, the decision on the data model is often driven by data and
software availability, or the familiarity with the approach (Costanza et
al. 2019). For example, several early landscape ecology studies were
based on polygonal vector data that was converted to raster data. Yet,
nowadays, vector data is increasingly available, and most geospatial
software supports it.

@@

@@

The open-source, script-based R programming languages (R Core Team 2024)
was first released in 1995 (Smith 2016) with a focus on statistical
computing and data visualization. However, due to its extensions (also
called packages), the language is popular in many academic fields,
including ecology (Lai et al. 2019; Lortie et al. 2020), and it is
suitable for spatial data analysis and landscape ecology (e.g., Bivand
2020; Hesselbarth et al. 2021). Fundamental packages for spatial data
analysis include \emph{sf} for simple-features vector data (Pebesma
2018), \emph{terra} for raster data (Hijmans 2024), \emph{stars} for
spatiotemporal arrays, raster, and vector data cubes (Pebesma and Bivand
2023), and \emph{spatstat} for point pattern data (Baddeley et al.
2015). Several packages allow to create maps, such as \emph{tmap}
(Tennekes 2018) or \emph{ggplot2} (Wickham 2016). For a comprehensive
overview of packages related to geographical spatial data, please see
the CRAN Task View (Bivand and Nowosad 2024).

R packages developed specifically for spatial pattern analysis include,
besides others, \emph{landscapemetrics} (Hesselbarth et al. 2019),
\emph{vectormetrics} {[}still being developed; Matuszek et al.
(2024){]}, \emph{motif} (Nowosad 2021), \emph{rasterdiv} (Rocchini et
al. 2021), or \emph{geodiv} (Smith et al. 2021). Packages with a focus
on scale include e.g., \emph{multilandr} (Huais 2024) or
\emph{grainchanger} (Graham 2024). Connectivity analysis can be
conducted using packages including \emph{lconnect} (Mestre and Silva
2023), or \emph{grainscape} (Chubaty et al. 2020). Furthermore, neutral
landscapes can be simulated using \emph{NLMR} (Sciaini et al. 2018), and
the \emph{graph4lg} (Savary et al. 2021) provides functions for
landscape genetics.

@@

\bibitem[\citeproctext]{ref-Bailey2007}
Bailey D, Billeter R, Aviron S, et al (2007) The influence of thematic
resolution on metric selection for biodiversity monitoring in
agricultural landscapes. Landscape Ecology 22:461--473

@@

\bibitem[\citeproctext]{ref-Bosch2019}
Bosch M (2019) PyLandStats: An open-source pythonic library to compute
landscape metrics. PLoS One 14:e0225734

\bibitem[\citeproctext]{ref-Buyantuyev2007}
Buyantuyev A, Wu J (2007) Effects of thematic resolution on landscape
pattern analysis. Landscape Ecology 22:7--13

@@

\bibitem[\citeproctext]{ref-Castilla2009}
Castilla G, Larkin K, Linke J, Hay GJ (2009) The impact of thematic
resolution on the patch-mosaic model of natural landscapes. Landscape
Ecology 24:15--23

@@

@@

\bibitem[\citeproctext]{ref-Frazier2014}
Frazier AE (2014) A new data aggregation technique to improve landscape
metric downscaling. Landscape ecology 29:1261--1276

\bibitem[\citeproctext]{ref-Frazier2016}
Frazier AE (2016) Surface metrics: Scaling relationships and downscaling
behavior. Landscape ecology 31:351--363

@@

@@

\bibitem[\citeproctext]{ref-Huang2006}
Huang C, Geiger E, Kupfer J (2006) Sensitivity of landscape metrics to
classification scheme. International Journal of Remote Sensing
27:2927--2948

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-Wu1999}
Wu J (1999) Hierarchy and scaling: Extrapolating information along a
scaling ladder. Canadian journal of remote sensing 25:367--380

>>>>

Despite the extensive use of the landscape mosaic model and metrics
based on it, major limitations have been identified regarding both its
ability to quantify the spatial structure of landscapes and their
relationship to ecological processes (Gustafson 1998; Li and Wu 2004;
Dramstad 2009; Kupfer 2012; Lausch et al. 2015; Frazier and Kedron
2017). Here, we review four major issues of landscape metrics, namely
their sensitivity to both spatial scale and thematic resolution,
redundancy and interpretation (Dramstad 2009).

@@

Identifying scales at which the ecological processes of interest operate
is a critical first step towards establishing relationships between
pattern and process (Levin 1992; Wu 2007; Cushman and McGarigal 2008).
To that end, a key requirement is to understand the sensitivity of
landscape metrics to spatial scale in terms of both grain and extent.

A key concept to consider regarding grain size is the minimum mapping
unit (MMU), namely the smallest area entity that can be mapped and
analyzed as a discrete patch. Since the input maps should represent not
only the location of patches but also their shape, the grain size should
be several times smaller than the smallest patch (i.e., the MMU) in
order to quantify shapes. Even though the MMU is often assumed as a
natural characteristic of the dataset, it has an impact on the computed
landscape metrics, especially on the number of patches and their size
distribution (Saura 2002).

While the choice of grain is often constrained by the available data,
e.g., Landsat scenes at 30 m resolution, the selection of spatial extent
varies largely depending on the phenomenon of interest, e.g., from large
transects of forests in natural parks to small buffers along the
urban-rural gradient. The spatial extent and sampling strategy is often
subject to many arbitrary criteria that can largely influence results.
This issue has long been recognized in physical geography as the
modifiable areal unit problem {[}MAUP; Openshaw (1979); Openshaw
(1984){]}. Dealing with the MAUP in the context of landscape metrics
include the identification of individual entities that are ecologically
meaningful and not modifiable, the evaluation of ``optimal'' zoning
systems that maximize interzonal variations and minimize intrazonal
variation, and the analysis of the sensitivity of the zoning
configuration (Jelinski and Wu 1996; Duque et al. 2018; Fletcher Jr et
al. 2023).

The scale effects of landscape metrics can be assessed by computing
scalograms, i.e., the response curves of metrics to changing scales.
Some metrics related to the number of patches, edges show consistent
linear and power-law scaling relationships to changing grain, while
other metrics exhibit less predictable responses (Turner et al. 1989;
Saura and Martínez-Millán 2001; Wu et al. 2002; Wu 2004; Saura and
Castro 2007; Šı́mová and Gdulová 2012; Argañaraz and Entraigas 2014;
Frazier et al. 2021). Nonetheless, the effects of changing the sampling
extent can be less predictable than those of changing grain size (Wu
2004). For instance, in landscapes with small extent to grain ratio,
boundary effects are likely to dominate and truncate patches resulting
in unreliable metric computations. In order to address scale issues,
studies should explicitly consider the concept of scope, namely the
ratio of extent to resolution (Schneider 2001). This involves three
fundamental steps, i) considering scope guidelines when designing
experiments, ii) explicitly reporting grain and extent allowing
quantitative scope comparisons with other studies, iii) and comparisons
and replications based on scope (Frazier 2023).

@@

Many landscape ecology studies are based on the raster data model.
However, the decision on the data model is often driven by data and
software availability, or the familiarity with the approach (Costanza et
al. 2019). For example, many LULC maps are provided as raster as they
often related to underlying remote sensing products. Yet, nowadays,
vector data is increasingly available, and most geospatial software
supports it.

@@

\section{Connectivity}\label{connectivity}

Connectivity describes how movement among resources patches is
facilitated or hindered by landscape features such as types, amounts,
and arrangement of LULC (Taylor et al. 1993, 2006). While structural
connectivity focuses on physical and spatial relationships among
resource patches, functional connectivity also includes the movement or
flow of individuals and populations through the landscape (Taylor et al.
2006). Connectivity underpins ecosystem functionality and maintains
biodiversity, and plays a large role in many conservation actions
(Correa Ayram et al. 2016; Hilty et al. 2020). Thus, in the past
multiple approaches how to measure connectivity were developed (Simpkins
et al. 2018; Keeley et al. 2021).

\subsection{Landscape metrics}\label{landscape-metrics-1}

\begin{itemize}
\tightlist
\item
  Many traditional lsm measure connectivity, e.g, ENN
\item
  Recently, also focus in within-patch connectivity
  (https://doi.org/10.1007/s10980-019-00881-0)
\end{itemize}

\subsection{Resistance surfaces}\label{resistance-surfaces}

https://doi.org/10.1111/2041-210X.12750
https://doi.org/10.1007/s10980-019-00789-9
https://doi.org/10.1007/s10980-022-01504-x
https://doi.org/10.1007/s10980-022-01469-x

\subsection{Circuit theory-based}\label{circuit-theory-based}

https://circuitscape.org https://doi.org/10.1111/2041-210X.12689
https://doi.org/10.1016/j.ecolind.2016.03.032

\subsection{Least-cost paths (LCPs)}\label{least-cost-paths-lcps}

https://doi.org/10.1016/S0169-2046(02)00242-6
https://doi.org/10.1007/s40823-016-0006-9

\subsection{Graph theory}\label{graph-theory}

https://doi.org/10.1007/s10980-020-00998-7
https://doi.org/10.1007/s10980-020-01164-9

\subsection{Various}\label{various}

https://doi.org/10.1890/14-1690.1
https://doi.org/10.1016/j.ecolind.2016.03.032
https://doi.org/10.1007/s10980-019-00935-3
https://doi.org/10.1126/science.abg1780

@@

The open-source, script-based R programming languages (R Core Team 2024)
was first released in 1995 (Smith 2016) with a focus on statistical
computing and data visualization. However, due to its extensions (also
called packages), the language is popular in many academic fields,
including ecology (Lai et al. 2019; Lortie et al. 2020), and it is
suitable for spatial data analysis and landscape ecology (e.g., Bivand
2020; Hesselbarth et al. 2021). Fundamental packages for spatial data
analysis include \emph{sf} for simple-features vector data (Pebesma
2018), \emph{terra} for raster data (Hijmans 2024), \emph{stars} for
spatiotemporal arrays, raster, and vector data cubes (Pebesma and Bivand
2023), and \emph{spatstat} for point pattern data (Baddeley et al.
2015). Several packages allow to create maps, such as \emph{tmap}
(Tennekes 2018) or \emph{ggplot2} (Wickham 2016). For a comprehensive
overview of packages related to geographical spatial data, please see
the CRAN Task View (Bivand and Nowosad 2024). R packages developed
specifically for spatial pattern analysis include, besides others,
\emph{landscapemetrics} (Hesselbarth et al. 2019), \emph{vectormetrics}
{[}still being developed; Matuszek et al. (2024){]}, \emph{motif}
(Nowosad 2021), \emph{rasterdiv} (Rocchini et al. 2021), or
\emph{geodiv} (Smith et al. 2021). Packages with a focus on scale
include e.g., \emph{multilandr} (Huais 2024) or \emph{grainchanger}
(Graham 2024). Connectivity analysis can be conducted using packages
including \emph{lconnect} (Mestre and Silva 2023), or \emph{grainscape}
(Chubaty et al. 2020). Furthermore, neutral landscapes can be simulated
using \emph{NLMR} (Sciaini et al. 2018), and the \emph{graph4lg} (Savary
et al. 2021) provides functions for landscape genetics.

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-CorreaAyram2016}
Correa Ayram CA, Mendoza ME, Etter A, Salicrup DRP (2016) Habitat
connectivity in biodiversity conservation: A review of recent studies
and applications. Progress in Physical Geography: Earth and Environment
40:7--37. \url{https://doi.org/10.1177/0309133315598713}

@@

@@

\bibitem[\citeproctext]{ref-Hilty2020}
Hilty J, Worboys GL, Keeley A, et al (2020)
\href{https://doi.org/10.2305/IUCN.CH.2020.PAG.30.en}{Guidelines for
conserving connectivity through ecological networks and corridors}.
IUCN, International Union for Conservation of Nature

@@

@@

\bibitem[\citeproctext]{ref-Keeley2021}
Keeley ATH, Beier P, Jenness JS (2021) Connectivity metrics for
conservation planning and monitoring. Biological Conservation
255:109008. \url{https://doi.org/10.1016/j.biocon.2021.109008}

@@

\bibitem[\citeproctext]{ref-Simpkins2018}
Simpkins CE, Dennis TE, Etherington TR, Perry GLW (2018) Assessing the
performance of common landscape connectivity metrics using a virtual
ecologist approach. Ecological Modelling 367:13--23.
\url{https://doi.org/10.1016/j.ecolmodel.2017.11.001}

@@

\bibitem[\citeproctext]{ref-Taylor1993}
Taylor PD, Fahrig L, Henein K, Merriam G (1993) Connectivity is a vital
element of landscape structure. Oikos 68:571.
\url{https://doi.org/10.2307/3544927}

\bibitem[\citeproctext]{ref-Taylor2006}
Taylor PD, Fahrig L, With KA (2006)
\href{https://doi.org/10.1017/CBO9780511754821}{Landscape connectivity:
A return to the basics}. In: Crooks KR, Sanjayan M (eds) Connectivity
conservation. Cambridge University Press, Cambridge, pp 29--43

@@
